### 2023-06-08 paper/localpsf.tex
editing paper based on Noemi's comments. Found bad typo in H-matrix complexity estimate (said kh^2 N log N entries needed, should be kh N log N)

We present an efficient \emph{matrix-free} point spread function (PSF) method for approximating operators $\Aop:L^2(\Omega) \rightarrow L^2(\Omega)'$ that have locally supported non-negative integral kernels. Here $\Omega \subset \mathbb{R}^\gdim$ is a bounded domain, and $L^2(\Omega)'$ is the space of real-valued continuous linear functionals on $L^2(\Omega)$. We focus on approximating Hessians in optimization and inverse problems governed by partial differential equations (PDEs)~\cite{AlgerEtAl19,IsaacEtAl15}. Such operators also arise as certain Schur complements in Schur complement methods for solving partial differential equations (PDEs)~\cite{le1997non,saad1999distributed}, Poincare-Steklov operators in domain decomposition methods (e.g., Dirichlet-to-Neumann maps), covariance operators in spatial statistics~\cite{ChenStein21,GeogaEtAl20,LindgrenRueLindstrom11}, and blurring operators in imaging~\cite{DenisEtAl11,NagyOleary98}. Here ``matrix-free'' means that we may apply $\Aop$ and its transpose\footnote{Recall that $\Aop^T:L^2(\Omega)\rightarrow L^2(\Omega)'$ is the unique operator satisfying $\left(\Aop u\right)(w) = \left(\Aop^T w\right)(u)$ for all $u,w \in L^2(\Omega)$, where $\Aop u \in L^2(\Omega)'$ is the result of applying $\Aop$ to $u\in L^2(\Omega)$, and $\left(\Aop u\right)(w)$ is the result of applying that linear functional to $w \in L^2(\Omega)$, and similar for operations with $\Aop^T$. }, $\Aop^T$, to arbitrary functions,
\begin{equation}
\label{eq:eval_maps}
u \mapsto\Aop u \qquad \text{and} \qquad w \mapsto \Aop^T w,
\end{equation} 
via a black box computational procedure, 
but cannot easily access entries of $\Aop$'s integral kernel. Evaluating the maps in~\eqref{eq:eval_maps} may require, as a sub procedure, solving PDEs or other costly computations. By ``non-negative integral kernel,'' we mean that entries of the integral kernel are non-negative numbers; this is not the same as positive semi-definiteness of $\Aop$.

\begin{figure}
	\begin{subfigure}{0.32\textwidth}
		\centering
		\includegraphics[scale=0.34]{impulse_batch1.png}
	\end{subfigure}
	\begin{subfigure}{0.32\textwidth}
		\centering
		\includegraphics[scale=0.34]{impulse_batch2.png}
	\end{subfigure}
	\begin{subfigure}{0.32\textwidth}
		\centering
		\includegraphics[scale=0.34]{impulse_batch3.png}
	\end{subfigure}
	\caption{Batches, $\eta_b$, of normalized impulses, $\impulseresponse_{x}$, for the ice sheet inverse problem data misfit Gauss-Newton Hessian described in Section~\ref{sec:numerical_results}. 
%	Brightness scale differs between the subfigures. 
	Black stars are point source locations. Dashed gray ellipses are estimated impulse response support ellipsoids. The large circle is $\partial \Omega$. 
	}
	\label{fig:batches_intro}
\end{figure}

@@

\begin{figure}
	\begin{center}
		\includegraphics[scale=0.75]{mean_displacement_invariance.pdf}
	\end{center}
	\caption{$\Aop$ is locally mean displacement invariant if $\impulseresponse_{x}(y) \approx \impulseresponse_{x'}\left(y - \spatialmean(x) + \spatialmean(x')\right)$ when $x$ is close to $x'$. Here $\spatialmean(z)$ denotes the mean (center of mass) of $\phi_z$.}
	\label{fig:mean_displacement_invariance}
\end{figure}

The ability to rapidly approximate entries of $\Aop$'s integral kernel allows one to approximate discretized versions of $\Aop$ using the full arsenal of tools for matrix approximation that rely on fast access to matrix entries.
%one to approximate discretized matrix representations of $\Aop$ using methods that require fast access to arbitrary matrix entries. We 
We form a hierarchical matrix (H-matrix)~\cite{BormGrasedyckHackbusch03,Hackbusch99} approximation of a discretized version of $\Aop$. H-matrices are a matrix format in which the rows and columns of the matrix are re-ordered, then the matrix is recursively subdivided into blocks, in such a way that many off-diagonal blocks are low rank, even though the matrix as a whole may be high rank. 
%(see Figure~\ref{fig:hmatrix_intro} and Appendix~\ref{app:h_matrix}).
H-matrix methods permit us to perform matrix-vector products cheaply, and perform other useful linear algebra operations that cannot be done easily using the original operator. These operations include matrix-matrix addition, matrix-matrix multiplication, matrix factorization, and matrix inversion. 
%These H-matrix methods are fast and scale well to large problems. 
The work and memory required to perform these operations for an $\fedim \times \fedim$ H-matrix scales 
%as $O(\fedim \log(\fedim)^a)$, where $a \in \{0,1,2,3\}$. That is, the cost is \emph{nearly-linear}
\emph{nearly linearly} in $N$ (i.e., $o(N^{1+\epsilon})$ for any $\epsilon > 0$). The exact cost depends on the type of H-matrix used, the operation being performed, and the rank of the off-diagonal blocks~\cite{GrasedyckHackbusch03}. These operations can be used, for example, to solve Newton linear systems in PDE constrained optimization, sample from ill-conditioned posterior distributions in Bayesian inverse problems, and construct high rank Gaussian surrogate models that can be used for uncertainty quantification. 
%Classical H-matrix construction methods 
%(described in Appendix~\ref{sec:H_matrix_construction}) 
%require access to matrix entries of the matrix being approximated, and therefore cannot be used to efficiently form H-matrix approximations of operators that are only available matrix free.

@@

In distributed parameter inverse problems governed by PDEs, one seeks to infer an unknown spatially varying parameter field from limited observations of a state variable that depends on the parameter implicitly through the solution of a PDE. 
%Inverse problems are typically ill-posed, in the sense that many different parameter fields may yield predicted observations that agree with the actual observations to within the noise level. 
Conventionally, the inverse problem is formulated using either a deterministic framework~\cite{AsterBorchersThurber18,Vogel02},
%Guy Chavent, Nonlinear Least Squares for Inverse Problems, Springer, 2009.
or a Bayesian probabilistic framework~\cite{KaipioSomersalo05,Stuart10a,Tarantola05}. In the deterministic framework, one solves an optimization problem to find the parameter that best fits the observations, subject to appropriate regularization~\cite{EnglHankeNeubauer96}. In the probabilistic framework, Bayes' theorem combines the observations with prior information to form a posterior distribution over the space of all possible parameter fields, and computations are performed to extract statistical information about the parameter from this posterior. The Hessian of the objective function in the determinstic optimization problem and the Hessian of the negative log posterior in the Bayesian setting are equal or approximately equal to each other under typical noise, regularization, and prior models, so we refer to both of these Hessians as ``the Hessian.'' The Hessian consists of a data misfit term (the ``data misfit Hessian''), which depends on a discrepancy between the observations and the associated model predictions, and a regularization or prior term (the ``regularization Hessian'') which does not depend on the observations. For more details on the Hessian, see~\cite{Alger19,GhattasWillcox21,VillaPetraGhattas21}.

@@

Examples of inverse problems with high rank data misfit Hessians include advection dominated advection-diffusion inverse problems~\cite{AkcelikBirosDraganescuEtAl05}\cite[Chapter 5]{Flath13}, high frequency wave inverse problems~\cite{BuiEtAl13}, and more generally, all inverse problems in which the observations highly inform the parameter. The eigenvalues of the data misfit Hessian characterize the sensitivity of the predicted observations to perturbations of the parameter in the corresponding eigenvector directions,
%(see Figure~\ref{fig:hessian_eigenstructure})
so more informative data leads to larger eigenvalues and therefore a larger numerical rank~\cite{AlexanderianGloorGhattas16}\cite[Section 1.4 and Chapter 4]{Alger19}. Roughly speaking, the numerical rank of the data misfit Hessian is the dimension of the subspace of parameter space that is informed by the data. The numerical rank of the regularization preconditioned data misfit Hessian may be reduced by increasing the strength of the regularization, but this throws away useful information---components of the parameter that could be learned from components of the observations that have a large signal to noise ratio are instead reconstructed based on the regularization~\cite[Section 4]{AlgerEtAl17}\cite[Chapters 1 and 7]{Vogel02}. Hence, low rank approximation methods suffer from a predicament---if the data highly inform the parameter and the regularization is chosen appropriately, then a large number of operator applies are required to form an accurate approximation of the Hessian using low rank approximation methods. 
%The best case scenario from a scientific perspective (highly informative data) is the worst case scenario from a computational perspective (large computational cost)~\cite{AlexanderianGloorGhattas16,Alger19,AlgerEtAl17}. 
High rank Hessian approximation methods are needed.

@@

The action of $\Aop$ may be extended to distributions if $\Aker$ is sufficiently regular. 
%We derive the extension as follows. 
Let $\rho \in L^2(\Omega)'$, and let $\rho^* \in L^2(\Omega)$ denote the Riesz representative of $\rho$ with respect to the $L^2(\Omega)$ inner product. We have
\begin{subequations}
	\label{eq:extension_to_distributions}
	\begin{align}
		\left(\Aop \genericdistribution^*\right)(w) &= \int_\Omega \int_\Omega w(y) \Aker(y,x) \genericdistribution^*(x) \mathrm{d}x ~\mathrm{d}y \\
		&= \int_\Omega w(y) \int_\Omega \Aker(y,x) \genericdistribution^*(x) \mathrm{d}x ~\mathrm{d}y 
		= \int_\Omega w(y) \genericdistribution\left(\Aker(y,\cdot)\right) \mathrm{d}y, \label{eq:action_on_distribution}
	\end{align}
\end{subequations}
where $\Aker(y,\cdot)$ denotes the function $x \mapsto \Aker(y,x)$.
Now let $\mathcal{D}(\Omega) \subset L^2(\Omega)$ be a suitable space of test functions and let $\rho:\mathcal{D}(\Omega) \rightarrow \mathbb{R}$ be a distribution. In this case, the Riesz representative $\rho^*$ may not exist, so the derivation in~\eqref{eq:extension_to_distributions} is not valid. However, if $\Aker$ is sufficiently regular such that the function $y \mapsto \rho\left(\Aker(y,~\cdot~)\right)$ is well defined for almost all $y\in\Omega$, and if this function is in $L^2(\Omega)$, then the right hand side of~\eqref{eq:action_on_distribution} is well defined. Hence, we \emph{define} the action of $\Aop$ on the distribution $\rho$ to be the right hand side of~\eqref{eq:action_on_distribution}. We denote this action by ``$\Aop \genericdistribution^*$,'' even if $\rho^*$ does not exist.

@@

We focus on approximating operators that satisfy the following conditions: 
\begin{enumerate}
	\item The kernel $\Aker$ is sufficiently regular so that $\phi_x$ is well defined for all $x\in \Omega$.
	\item The supports of the impulse responses $\impulseresponse_x$ are contained in localized regions.
	\item The integral kernel is non-negative in the sense that $$\Aker(y,x) \ge 0$$ for all $(y,x) \in \Omega \times \Omega$.\footnote{Note that having a non-negative integral kernel is different from positive semi-definiteness. The operator $\Aop$ need not be positive semi-definite to use the proposed method, and positive semi-definite operators need not have a non-negative integral kernel.}
\end{enumerate}
The proposed method may still perform well if these conditions are relaxed slightly. It is fine if the support of $\impulseresponse_x$ is not perfectly contained in a localized region, as long as the bulk of the ``mass'' of $\impulseresponse_x$ is contained in a localized region. The integral kernel does not need to be non-negative for all pairs of points $(y,x) \in \Omega \times \Omega$, as long as it is non-negative for the vast majority of pairs of points $(y,x)$, and as long as the negative numbers are comparatively small. If these conditions are violated, the method will incur additional error. If these conditions are violated too much, the method may fail. 
%Not all operators of interest (and in particular not all Hessians) satisfy these conditions, but many operators of practical interest do satisfy these conditions, and our method is highly effective for approximating these operators.

@@

In computations, functions are discretized and replaced by finite dimensional vectors, and operators mapping between infinite dimensional spaces are replaced by operators mapping between finite dimensional spaces. In this paper we discretize using continuous Galerkin finite elements satisfying the Kronecker property (defined below). With minor modifications, the proposed method could be used with more general finite element methods, or other discretization schemes such as finite differences or finite volumes.

@@

After discretization, the operator $\Aop:L^2(\Omega) \rightarrow L^2(\Omega)'$ is replaced by an operator $A_h:V_h \rightarrow V_h'$, which becomes an operator
\begin{equation*}
\mathbf{A}:\mathbb{R}^\fedim_\massmatrix \rightarrow \mathbb{R}^\fedim_{\massmatrix^{-1}}
\end{equation*}
under the isomorphism discussed above. The proposed method is agnostic to the computational procedure for approximating $\Aop$ with $\mathbf{A}$. What is important is that we do not have direct access to matrix entries $\mathbf{A}_{ij}$. Rather, we have a computational procedure that allows us to compute matrix-vector products $\mathbf{u}\mapsto \mathbf{A}\mathbf{u}$ and $\mathbf{w}\mapsto \mathbf{A}^T\mathbf{w}$, and computing these matrix-vector products is costly.
Of course, matrix entries can be computed via matrix-vector products as $\mathbf{A}_{ij} = \left(\mathbf{A}\mathbf{e}_j\right)_i$,
where $\mathbf{e}_j=(0,\dots,0,1,0,\dots,0)^T$ is the length $\fedim$ unit vector with one in the $j$th coordinate and zeros elsewhere. But computing the matrix-vector product $\mathbf{e}_j \mapsto \mathbf{A}\mathbf{e}_j$ is costly, and therefore wasteful if we do not use other matrix entries in the $j$th column of $\mathbf{A}$. Hence, methods for approximating $\mathbf{A}$ are computationally intractable if they require accessing scattered matrix entries from many different rows and columns of $\mathbf{A}$.

@@

When $\Aop$ is symmetric positive semi-definite, $\mathbf{A}_H$ may be non-symmetric and indefinite due to errors in the approximation. In this case, one may optionally symmetrize $\mathbf{A}_H$, then modify it via low rank updates to remove erroneous negative eigenvalues (Section \ref{sec:flip_eigs}). The complete algorithm for constructing $\mathbf{A}_H$ is shown in Algorithm~\ref{alg:construct_Atilde}. The cost is discussed in Section~\ref{sec:overall_cost}.

@@

			Compute $\combresponse_k$ by applying $\Aop$ to the Dirac comb for $\pointbatch_k$ (Section~\ref{sec:get_impulse_response})

@@

		Form H-matrix approximation $\boldsymbol{\Aker}_H$ of integral kernel (Section~\ref{sec:h_matrix_construction_short})

@@

We compute impulse responses $\impulseresponse_{x_i}$ in batches by applying $\Aop$ to Dirac combs. The Dirac comb, $\diraccomb_k$, associated with a batch of sample points, $\pointbatch_k$, is the following weighted sum of Dirac distributions (point sources) centered at the points $x_i \in \pointbatch_k$:
\begin{equation*}
	\diraccomb_k := \sum_{x_i \in \pointbatch_k} \delta_{x_i} / \spatialvol(x_i).
\end{equation*}
We compute the \emph{impulse batch}, $\eta_k$, by applying $\Aop$ to the Dirac comb:
\begin{equation}
	\label{eq:dirac_comb_H_action}
	\combresponse_k := \left(\Aop \diraccomb_k^*\right)^*
	=\sum_{x_i \in \pointbatch_k} \impulseresponse_{x_i} / \spatialvol(x_i).
\end{equation}
The last equality in~\eqref{eq:dirac_comb_H_action} follows from linearity and the definition of $\impulseresponse_{x_i}$ in~\eqref{eq:impulse_response_delta_action}. Since the points $x_i$ are chosen so that the ellipsoid $E_{x_i}$ that (approximately) supports $\impulseresponse_i$, and the ellipsoid $E_{x_j}$ that (approximately) supports $\impulseresponse_j$ do not overlap when $i \neq j$, we have (approximately)
\begin{equation}
\label{eq:varphi_eval}
	\impulseresponse_{x_i}(z) =
	\begin{cases}
		\combresponse_k(z) \spatialvol(x_i), & z \in E_{x_i} \\
		0, & \text{otherwise}
	\end{cases}
\end{equation}
for all $x_i \in \pointbatch_k$. By performing one matrix-vector product, $\xi_k \mapsto \left(\Aop \diraccomb_k^*\right)^*$, we recover $\impulseresponse_{x_i}$ for every point $x_i \in \pointbatch_k$.

@@

We form an H-matrix approximation $\mathbf{A}_H \approx \mathbf{A}$ by forming an H-matrix representation $\mathbf{\Aker}_H$ of $\mathbf{\Aker}$ then multiplying $\mathbf{\Aker}$ with mass matrices $\massmatrix$ per~\eqref{eq:boldA} to form
$
\mathbf{A}_H = \massmatrix \mathbf{\Aker}_H \massmatrix.
$
In the numerical results, $\massmatrix$ is a lumped mass matrix and is therefore diagonal, so these matrix-matrix multiplications are trivial. If a non-diagonal mass matrix is used, one may form a H-matrix representation of the mass matrix, then perform the matrix-matrix multiplications in \eqref{eq:boldA} using H-matrix methods.
%Once $\mathbf{A}_H$ is formed, useful matrix operations such as matrix-vector products, matrix-matrix addition, matrix-matrix multiplication, and matrix inversion may be performed using fast scalable H-matrix methods~\cite{BORMHACKBUSCHBOOK}. 
We use H1 matrices in the numerical results, but any other H-matrix format could be used instead. For more details on H-matrices, see~\cite{Hackbusch15}.

>>>>

We present an efficient \emph{matrix-free} point spread function (PSF) method for approximating operators $\Aop:L^2(\Omega) \rightarrow L^2(\Omega)'$ that have locally supported non-negative integral kernels. Here $\Omega \subset \mathbb{R}^\gdim$ is a bounded domain, and $L^2(\Omega)'$ is the space of real-valued continuous linear functionals on $L^2(\Omega)$. Such operators appear, for instance, as Hessians in optimization and inverse problems governed by partial differential equations (PDEs)~\cite{AntilEtAl18,BorziSchulz11}, Schur complements in Schur complement methods for solving partial differential equations (PDEs)~\cite{le1997non,saad1999distributed}, Poincare-Steklov operators in domain decomposition methods (e.g., Dirichlet-to-Neumann maps), covariance operators in spatial statistics~\cite{ChenStein21,GeogaEtAl20,LindgrenRueLindstrom11}, and blurring operators in imaging~\cite{DenisEtAl11,NagyOleary98}. Here ``matrix-free'' means that we may apply $\Aop$ and its transpose\footnote{Recall that $\Aop^T:L^2(\Omega)\rightarrow L^2(\Omega)'$ is the unique operator satisfying $\left(\Aop u\right)(w) = \left(\Aop^T w\right)(u)$ for all $u,w \in L^2(\Omega)$, where $\Aop u \in L^2(\Omega)'$ is the result of applying $\Aop$ to $u\in L^2(\Omega)$, and $\left(\Aop u\right)(w)$ is the result of applying that linear functional to $w \in L^2(\Omega)$, and similar for operations with $\Aop^T$. }, $\Aop^T$, to arbitrary functions,
\begin{equation}
\label{eq:eval_maps}
u \mapsto\Aop u \qquad \text{and} \qquad w \mapsto \Aop^T w,
\end{equation} 
via a black box computational procedure, 
but cannot easily access entries of $\Aop$'s integral kernel. Evaluating the maps in~\eqref{eq:eval_maps} may require solving a sub problem that involves PDEs, or performing other costly computations. By ``non-negative integral kernel,'' we mean that entries of the integral kernel are non-negative numbers; this is not the same as positive semi-definiteness of $\Aop$.

\begin{figure}
	\begin{subfigure}{0.32\textwidth}
		\centering
		\includegraphics[scale=0.34]{impulse_batch1.png}
	\end{subfigure}
	\begin{subfigure}{0.32\textwidth}
		\centering
		\includegraphics[scale=0.34]{impulse_batch2.png}
	\end{subfigure}
	\begin{subfigure}{0.32\textwidth}
		\centering
		\includegraphics[scale=0.34]{impulse_batch3.png}
	\end{subfigure}
	\caption{Three batches, $\eta_b$, of normalized impulse responses, $\impulseresponse_{x}$, that arise from applying $\Aop$ to weighted sums of scattered point sources (see Section \ref{sec:get_impulse_response}). Here $\Aop$ is the ice sheet inverse problem data misfit Gauss-Newton Hessian described in Section~\ref{sec:numerical_results}. 
%	Brightness scale differs between the subfigures. 
	Black stars are point source locations. Shading shows the magnitude of the normalized impulse responses (darker means larger function values). Dashed gray ellipses are estimated impulse response support ellipsoids based on the moment method in Section \ref{sec:intromoments}. The large circle is $\partial \Omega$. 
	}
	\label{fig:batches_intro}
\end{figure}

@@

\begin{figure}
	\begin{center}
		\includegraphics[scale=0.75]{mean_displacement_invariance.pdf}
	\end{center}
	\caption{This figure illustrates impulse responses, $\impulseresponse_x$ and $\impulseresponse_{x'}$, that arise from applying $\Aop$ to point sources at two points, $x$ and $x'$. The operator $\Aop$ is locally mean displacement invariant (Section \ref{sec:local_mean_displacement_invariance}) if $\impulseresponse_{x}(y) \approx \impulseresponse_{x'}\left(y - \spatialmean(x) + \spatialmean(x')\right)$ when $x$ is close to $x'$. Here $\spatialmean(z)$ denotes the mean (center of mass) of $\phi_z$.}
	\label{fig:mean_displacement_invariance}
\end{figure}

The ability to rapidly approximate entries of $\Aop$'s integral kernel allows one to approximate discretized versions of $\Aop$ using the full arsenal of tools for matrix approximation that rely on fast access to matrix entries.
%one to approximate discretized matrix representations of $\Aop$ using methods that require fast access to arbitrary matrix entries. We 
In this work, we form a hierarchical matrix (H-matrix)~\cite{BormGrasedyckHackbusch03,Hackbusch99} approximation of a discretized version of $\Aop$. H-matrices are a matrix format in which the rows and columns of the matrix are re-ordered, then the matrix is recursively subdivided into blocks, in such a way that many off-diagonal blocks are low rank, even though the matrix as a whole may be high rank. 
%(see Figure~\ref{fig:hmatrix_intro} and Appendix~\ref{app:h_matrix}).
H-matrix methods permit us to perform matrix-vector products cheaply, and perform other useful linear algebra operations that cannot be done easily using the original operator. These operations include matrix-matrix addition, matrix-matrix multiplication, matrix factorization, and matrix inversion. 
%These H-matrix methods are fast and scale well to large problems. 
The work and memory required to perform these operations for an $\fedim \times \fedim$ H-matrix scales 
%as $O(\fedim \log(\fedim)^a)$, where $a \in \{0,1,2,3\}$. That is, the cost is \emph{nearly-linear}
\emph{nearly linearly} in $N$ (i.e., $o(N^{1+\epsilon})$ for any $\epsilon > 0$). The exact cost depends on the type of H-matrix used, the operation being performed, and the rank of the off-diagonal blocks~\cite{GrasedyckHackbusch03}. 
%These operations can be used, for example, to solve Newton linear systems in PDE constrained optimization, sample from ill-conditioned posterior distributions in Bayesian inverse problems, and construct high rank Gaussian surrogate models that can be used for uncertainty quantification.

%Classical H-matrix construction methods 
%(described in Appendix~\ref{sec:H_matrix_construction}) 
%require access to matrix entries of the matrix being approximated, and therefore cannot be used to efficiently form H-matrix approximations of operators that are only available matrix free.

@@

In distributed parameter inverse problems governed by PDEs, one seeks to infer an unknown spatially varying parameter field from limited observations of a state variable that depends on the parameter implicitly through the solution of a PDE. 
%Inverse problems are typically ill-posed, in the sense that many different parameter fields may yield predicted observations that agree with the actual observations to within the noise level. 
Conventionally, the inverse problem is formulated using either a deterministic framework~\cite{AsterBorchersThurber18,BorziSchulz11,Vogel02},
%Guy Chavent, Nonlinear Least Squares for Inverse Problems, Springer, 2009.
or a Bayesian probabilistic framework~\cite{KaipioSomersalo05,Stuart10a,Tarantola05}. In the deterministic framework, one solves an optimization problem to find the parameter that best fits the observations, subject to appropriate regularization~\cite{EnglHankeNeubauer96,Vogel02}. In the probabilistic framework, Bayes' theorem combines the observations with prior information to form a posterior distribution over the space of all possible parameter fields, and computations are performed to extract statistical information about the parameter from this posterior. The Hessian of the objective function in the determinstic optimization problem and the Hessian of the negative log posterior in the Bayesian setting are equal or approximately equal to each other under typical noise, regularization, and prior models, so we refer to both of these Hessians as ``the Hessian.'' The Hessian consists of a data misfit term (the \emph{data misfit Hessian}), which depends on a discrepancy between the observations and the associated model predictions, and a regularization or prior term (the ``regularization Hessian'') which does not depend on the observations. For more details on the Hessian, see~\cite{Alger19,GhattasWillcox21,VillaPetraGhattas21}.

@@

Examples of inverse problems with high rank data misfit Hessians include large-scale ice sheet inverse problems \cite{HartlandEtAl23,IsaacEtAl15}, advection dominated advection-diffusion inverse problems~\cite{AkcelikBirosDraganescuEtAl05}\cite[Chapter 5]{Flath13}, high frequency wave inverse problems~\cite{BuiEtAl13}, and more generally, all inverse problems in which the observations highly inform the parameter. The eigenvalues of the data misfit Hessian characterize the sensitivity of the predicted observations to perturbations of the parameter in the corresponding eigenvector directions,
%(see Figure~\ref{fig:hessian_eigenstructure})
hence more informative data leads to larger eigenvalues and therefore a larger numerical rank~\cite{AlexanderianGloorGhattas16}\cite[Section 1.4 and Chapter 4]{Alger19}. Roughly speaking, the numerical rank of the data misfit Hessian is the dimension of the subspace of parameter space that is informed by the data. The numerical rank of the regularization preconditioned data misfit Hessian may be reduced by increasing the strength of the regularization, but this throws away useful information---components of the parameter that could be learned from components of the observations that have a large signal to noise ratio are instead reconstructed based on the regularization~\cite[Section 4]{AlgerEtAl17}\cite[Chapters 1 and 7]{Vogel02}. Hence, low rank approximation methods suffer from a predicament---if the data highly inform the parameter and the regularization is chosen appropriately, then a large number of operator applies are required to form an accurate approximation of the Hessian using low rank approximation methods. 
%The best case scenario from a scientific perspective (highly informative data) is the worst case scenario from a computational perspective (large computational cost)~\cite{AlexanderianGloorGhattas16,Alger19,AlgerEtAl17}. 
High rank Hessian approximation methods are needed.

@@

The action of $\Aop$ may be extended to distributions if $\Aker$ is sufficiently regular. 
%We derive the extension as follows. 
Let $\rho \in L^2(\Omega)'$, and let $\rho^* \in L^2(\Omega)$ denote the Riesz representative of $\rho$ with respect to the $L^2(\Omega)$ inner product. We have
\begin{subequations}
	\label{eq:extension_to_distributions}
	\begin{align}
		\left(\Aop \genericdistribution^*\right)(w) &= \int_\Omega \int_\Omega w(y) \Aker(y,x) \genericdistribution^*(x) \mathrm{d}x ~\mathrm{d}y \\
		&= \int_\Omega w(y) \int_\Omega \Aker(y,x) \genericdistribution^*(x) \mathrm{d}x ~\mathrm{d}y 
		= \int_\Omega w(y) \genericdistribution\left(\Aker(y,\cdot)\right) \mathrm{d}y, \label{eq:action_on_distribution}
	\end{align}
\end{subequations}
where $\Aker(y,\cdot)$ denotes the function $x \mapsto \Aker(y,x)$.
Now let $\mathcal{D}(\Omega) \subset L^2(\Omega)$ be a suitable space of test functions and let $\rho:\mathcal{D}(\Omega) \rightarrow \mathbb{R}$ be a distribution. In this case, the Riesz representative $\rho^*$ may not exist, so the derivation in~\eqref{eq:extension_to_distributions} is not valid. However, if $\Aker$ is sufficiently regular such that the function $y \mapsto \rho\left(\Aker(y,~\cdot~)\right)$ is well-defined for almost all $y\in\Omega$, and if this function is in $L^2(\Omega)$, then the right hand side of~\eqref{eq:action_on_distribution} is well-defined. Hence, we \emph{define} the action of $\Aop$ on the distribution $\rho$ to be the right hand side of~\eqref{eq:action_on_distribution}. We denote this action by ``$\Aop \genericdistribution^*$,'' even if $\rho^*$ does not exist.

@@

We focus on approximating operators that satisfy the following conditions: 
\begin{enumerate}
	\item The kernel $\Aker$ is sufficiently regular so that $\phi_x$ is well-defined for all $x\in \Omega$.
	\item The supports of the impulse responses $\impulseresponse_x$ are contained in localized regions.
	\item The integral kernel is non-negative in the sense that $$\Aker(y,x) \ge 0$$ for all $(y,x) \in \Omega \times \Omega$.\footnote{Note that having a non-negative integral kernel is different from positive semi-definiteness. The operator $\Aop$ need not be positive semi-definite to use the proposed method, and positive semi-definite operators need not have a non-negative integral kernel.}
\end{enumerate}
The proposed method may still perform well if these conditions are relaxed slightly. It is acceptable if the support of $\impulseresponse_x$ is not perfectly contained in a localized region, so long as the bulk of the ``mass'' of $\impulseresponse_x$ is contained in a localized region. The integral kernel does not need to be non-negative for all pairs of points $(y,x) \in \Omega \times \Omega$, as long as it is non-negative for the vast majority of pairs of points $(y,x)$, and as long as the negative numbers are comparatively small. If these conditions are violated, the method will incur additional error, and depending on the severity of the violation, the proposed method may fail.
%Not all operators of interest (and in particular not all Hessians) satisfy these conditions, but many operators of practical interest do satisfy these conditions, and our method is highly effective for approximating these operators.

@@

In computations, functions are discretized and replaced by finite-dimensional vectors, and operators mapping between infinite-dimensional spaces are replaced by operators mapping between finite-dimensional spaces. In this paper we discretize using continuous Galerkin finite elements satisfying the Kronecker property (defined below). With minor modifications, the proposed method could be used with more general finite element methods, or other discretization schemes such as finite differences or finite volumes.

@@

After discretization, the operator $\Aop:L^2(\Omega) \rightarrow L^2(\Omega)'$ is replaced by an operator $A_h:V_h \rightarrow V_h'$, which becomes an operator
\begin{equation*}
\mathbf{A}:\mathbb{R}^\fedim_\massmatrix \rightarrow \mathbb{R}^\fedim_{\massmatrix^{-1}}
\end{equation*}
under the isomorphism discussed above. The proposed method is agnostic to the computational procedure for approximating $\Aop$ with $\mathbf{A}$. What is important is that we do not have direct access to matrix entries $\mathbf{A}_{ij}$. Rather, we have a computational procedure that allows us to compute matrix-vector products $\mathbf{u}\mapsto \mathbf{A}\mathbf{u}$ and $\mathbf{w}\mapsto \mathbf{A}^T\mathbf{w}$, and computing these matrix-vector products is costly. The proposed method is designed to mitigate this cost by performing as few of these matrix-vector products as possible.
Of course, matrix entries can be computed via matrix-vector products as $\mathbf{A}_{ij} = \left(\mathbf{A}\mathbf{e}_j\right)_i$,
where $\mathbf{e}_j=(0,\dots,0,1,0,\dots,0)^T$ is the length $\fedim$ unit vector with one in the $j$th coordinate and zeros elsewhere. But computing the matrix-vector product $\mathbf{e}_j \mapsto \mathbf{A}\mathbf{e}_j$ is costly, and therefore wasteful if we do not use other matrix entries in the $j$th column of $\mathbf{A}$. Hence, methods for approximating $\mathbf{A}$ are computationally intractable if they require accessing scattered matrix entries from many different rows and columns of $\mathbf{A}$.

@@

When $\Aop$ is symmetric positive semi-definite, $\mathbf{A}_H$ may be non-symmetric and indefinite due to errors in the approximation. In this case, one may optionally symmetrize $\mathbf{A}_H$, then modify it via low rank updates to remove erroneous negative eigenvalues (Section \ref{sec:flip_eigs}). The complete algorithm for constructing $\mathbf{A}_H$ is shown in Algorithm~\ref{alg:construct_Atilde}. The computational cost is discussed in Section~\ref{sec:overall_cost}.

@@

			Compute impulse response batch $\combresponse_k$ by applying $\Aop$ to the Dirac comb for $\pointbatch_k$ (Section~\ref{sec:get_impulse_response})

@@

		Form H-matrix approximation $\boldsymbol{\Aker}_H$ of integral kernel (Sections~\ref{sec:approximate_kernel_entries} and \ref{sec:h_matrix_construction_short})

@@

We compute impulse responses $\impulseresponse_{x_i}$ in batches by applying $\Aop$ to Dirac combs. The Dirac comb, $\diraccomb_k$, associated with a batch of sample points, $\pointbatch_k$, is the following weighted sum of Dirac distributions (point sources) centered at the points $x_i \in \pointbatch_k$:
\begin{equation*}
	\diraccomb_k := \sum_{x_i \in \pointbatch_k} \delta_{x_i} / \spatialvol(x_i).
\end{equation*}
We compute the \emph{impulse response batch}, $\eta_k$, by applying $\Aop$ to the Dirac comb:
\begin{equation}
	\label{eq:dirac_comb_H_action}
	\combresponse_k := \left(\Aop \diraccomb_k^*\right)^*
	=\sum_{x_i \in \pointbatch_k} \impulseresponse_{x_i} / \spatialvol(x_i).
\end{equation}
The last equality in~\eqref{eq:dirac_comb_H_action} follows from linearity and the definition of $\impulseresponse_{x_i}$ in~\eqref{eq:impulse_response_delta_action}. Since the points $x_i$ are chosen so that the ellipsoid $E_{x_i}$ that (approximately) supports $\impulseresponse_i$, and the ellipsoid $E_{x_j}$ that (approximately) supports $\impulseresponse_j$ do not overlap when $i \neq j$, we have (approximately)
\begin{equation}
\label{eq:varphi_eval}
	\impulseresponse_{x_i}(z) =
	\begin{cases}
		\combresponse_k(z) \spatialvol(x_i), & z \in E_{x_i} \\
		0, & \text{otherwise}
	\end{cases}
\end{equation}
for all $x_i \in \pointbatch_k$. By performing one matrix-vector product, $\xi_k \mapsto \left(\Aop \diraccomb_k^*\right)^*$, we recover $\impulseresponse_{x_i}$ for every point $x_i \in \pointbatch_k$.

@@

We form an H-matrix approximation $\mathbf{A}_H \approx \mathbf{A}$ by forming an H-matrix representation $\mathbf{\Aker}_H$ of $\mathbf{\Aker}$ then multiplying $\mathbf{\Aker}$ with mass matrices $\massmatrix$ per~\eqref{eq:boldA} to form
$
\mathbf{A}_H = \massmatrix \mathbf{\Aker}_H \massmatrix.
$
Here we use a diagonal lumped mass matrix, so these matrix-matrix multiplications are trivial. If a non-diagonal mass matrix is used, one may form a H-matrix representation of the mass matrix, then perform the matrix-matrix multiplications in \eqref{eq:boldA} using H-matrix methods.
%Once $\mathbf{A}_H$ is formed, useful matrix operations such as matrix-vector products, matrix-matrix addition, matrix-matrix multiplication, and matrix inversion may be performed using fast scalable H-matrix methods~\cite{BORMHACKBUSCHBOOK}. 
We use H1 matrices in the numerical results, but any other H-matrix format could be used instead. For more details on H-matrices, see~\cite{Hackbusch15}.
